\section{introduction}

% circt 似乎也可以支持sv了https://discourse.llvm.org/t/is-circt-able-to-analyze-verilog/66281/22
CIRCT~\ToCite{} (Circuit IR Compilers and Tools) has emerged as a significant infrastructure in the realm of hardware compilers and tools, influenced by the success of LLVM~\ToCite{} in software. It utilizes MLIR~\ToCite{} (Multi-Level Intermediate Representation) dialects to offer a flexible and interoperable framework, which has been increasingly adopted by new HDLs (Hardware Description Languages), including Chisel~\ToCite{}, Magma~\ToCite{}, and Moore~\ToCite{}, for compiler development.

Unfortunately, CIRCT, devoid of formal semantics, introduces uncertainty and ambiguity, challenging its reliability in hardware design processes. This absence undermines the correctness and verifiability of the entire system, from input verification to output validation. As a result, the trustworthiness of hardware compilers, HDLs, and their dependent designs is compromised, marking a critical area for improvement.

% 相关工作融合在challenge中
\smallskip
\textbf{Goal and Challenge}. Accordingly, our goal is to establish formal semantics for CIRCT, thereby laying a rigorous and reliable foundation for contemporary hardware design. This endeavor is not only crucial for CIRCT's advancement but also brings forth a spectrum of challenges that must be meticulously addressed.

\ToStartMyDay{Do the research on the related work.}


% 或许先讲目前的主要工作是怎样的，然后CIRCT project有什么全新的需求，这种challenge使得我们需要考虑一种novel的方式以应对这种问题。
% 有做的好的，有做的不好的。反正都写上去。有些好的，就可以在下面写solution的时候把insight放进去。
\begin{enumerate}
    \item \ToOpt{\textit{Intrinsic Extensibility} poses the first challenge in formalizing CIRCT. Rooted in the MLIR framework, CIRCT is capable of incorporating various intermediate representations through MLIR's extensible architecture, particularly through MLIR dialects. These dialects, revolving around a unifying ``operation'' concept, allow for a broad spectrum of semantics to be captured, diverging from the fixed constructs of traditional languages such as statements and instructions. This inherent flexibility, while a strength, also put forward more requirements for the formalization, including unified formalizations to one-to-one map to all the unified concepts in CIRCT, keeping each dialect formalization independently definable and exist, as well as their compatibility and composability with each other. Not only that, CIRCT, a framework for hardware, distinct itself from LLVM and MLIR's initial design objectives, further complicating the formalization.}
    
    \textit{Existing Efforts}. \ToPreWork{1. Research on semantics composability. 2. Summary them in Related Work.} \ToWrite{(1)before don't care, since no need. recent work on compatibility since llvm, mlir (2)composability: game semantics \& effect handler. more? (3) work on mlir/llvm. (4) what's the gap? theoretically? Technology? In terms of implementation?} \ToResearch{Game Semantics}. \ToResearch{Effect Handler} \ToCite{Reasoning about MLIR Semantics through Effects and Handlers} \ToRead{Modular, compositional, and executable formal semantics for LLVM IR} \ToRead{Interaction trees: representing recursive and impure programs in Coq} \ToWrite{Ours: 1. Extensibility. 2. Hardware.} % 由于PL的发展，近期对于可组合性的谈论不断增多。Existing work 可以不考虑组合性问题（虽然有工作做），因为他们需要描述的语言可能是成熟且确定的，并且仅仅是一个语言，不需要去考虑可组合性问题。然而。甚至目前的一点点在MLIR形式语义上的工作都没有考虑到可组合性的问题。或者说没有把可组合性当作一个专门需要讨论的问题。
    \item \DateMark{2024-01-08}{To Write the following} \textit{Accessible Executability.}  Different from representations for software which are initially designed for execution, the representations for hardware are designed for circuit synthesis and description. Even ISAs like x86 or ARM are also designed for execution since they are compiled from the software programs and just requirements for the hardware. Thus, the semantics of hardware representations can be a translation from HDLs to circuit descriptions, which makes it hard to validate the correctness of the semantics. 
    
    \textit{Existing Efforts}. \ToResearch{Like Yosys... Recent works notice the importance of hardware simulation, including ... Some ... Some ... However, existing works are not keen on the accessibility of the formal semantics, which adapts to hardware designers' knowledge domain and is crucial for the adoption of the formal semantics.}
    % The existing works on formalizing hardware representations also meet this challenge, and they ??? % 时序特征的处理 % 仿真/可执行方面的处理。（或许需要读一下verilog形式语义方面的工作，特别是essense of verilog） % 然而，他们。。。。 % 我们希望探索和已有simulator近似的接口，将形式语义和一般人/硬件开发人员熟悉的语言结合，使形式语义不仅仅可以作为一个规范，而十分容易获取。 % 此外，与他们的目标语言不同。CIRCT中每一个dialect无法单独构成硬件的描述，而是需要组合起来。每个硬件描述仅仅负责描述硬件的一个方面。 % 硬件语言特征和软件语言特征的区别。时序。同样用K描述他的语义的时候也是有区别的。最后做仿真的时候也是有区别的。 % 时序这一方面是不一样的，这些语义重点去写。 % 用K描述硬件的语义和软件的语义是有一些差别的。描述的是电路的结构，是不会执行的，需要一些激励才会执行。在做仿真的时候，怎么去做硬件的激励和仿真？需要一个很好的interface。
     % C的符号执行和gcc还是比较像的。硬件的符号执行可能会有一些其他的问题。写越底层的程序，性能是一个很大的问题。
     % It necessitates altering the Kore language, which can be challenging to understand and modify, to enable continuous stimulation. Furthermore, its output does not comprise VCD files.
    \item \textit{Early Stage.} CIRCT is a new project and is still in its infancy. Although it is already used by some novel languages and maintained by the LLVM community, the flaws in the documentation and the modification of the infrastructure are inevitable. \ToThink{This not only requires the extensibility but also requires an essence of CIRCT as a stable and reliable foundation.} 
    
    \textit{Existing Efforts}. \ToResearch{CIRCT does have a core dialect set inside of its documentation, while as a baby project even without a research paper, the described core dialect set is not the real essence of it. (1) Existing work on formal semantics is basically working on mature languages that are more mature in terms of documentation, communities, etc. (2) (this maybe set in insight) Recent works ($\lambda_x$) present xxx, which is a good start. However, it is not enough for CIRCT.}
    \item \textit{Ambiguous Documentation.} The flaws in the language reference are not the only problem. Although the CIRCT project continues the LLVM/MLIR project's emphasis on documentation and maintains the same documentation standards as MLIR. As a usability-first compiler project, they write documentation for the people who want to contribute to it, rather than who want to use the languages. So the CIRCT dialects explanation is ambiguous and not enough from the perspective of formalization. For example, the \textit{hw.module} operation even misses the syntax and semantics explanation. This requires the formal semantics to be able to fill the gaps in the documentation. % 如果仅仅是early stage似乎还没有那么难，但是充满不确定性的文档带来了更大的难度。即便MLIR已经定义的标准的文档格式，和ref生成手段，但是。。。
    
    \textit{Existing Efforts}. \ToResearch{Almost all the language formalization faces this challenging ambiguity. This became particularly prominent after LLVM designed IR (intermediate representations) for compilation purposes, leading to a kind of transparent language for normal programmers. This trend has continued to increase with the birth of languages such as Rust, and thus MLIR has emerged. This trend of separation is such that we currently don't even have a formal semantics that claims to formalize MLIR. only set of MLIR dialects or Theoretical Framework for MLIR. As a hardware compilation architecture, CIRCT further extends its complexity. This is due to the dual design goals of simulation/synthesis, which can be seen from the characteristics of HDLs such as Verilog.}
    \item \textit{Comprehensive Validation} is inseparable from a well-designed test set. The existing test in CIRCT is mainly focused on transformation and compilation, with few semantic targeted tests in the CIRCT test suite. \ToThink{Moreover, the existing tests from simulators or compilers for other languages like Verilog may not cover the features of CIRCT's essence, because of the language differences. Despite the correctness, it is also challenging to validate the completeness of the essence wether it is indeed sufficient to describe actual circuits.}
    
    
    \textit{Existing Efforts}. \ToResearch{Other formal semantics work. especially K.} % 目前大部分都是Verilog的测试用例， % 并不完全一样。
\end{enumerate}


% work on 语义可拓展性， 语义可组合性。 Game Semantics

Thus, we can say that the hardware representations are not executable and need a simulator to stimulate them. This leads us to think about what the semantics of hardware representations look like. First of all, it should be executable and easy to use, like normal simulators, for validation. And it's vague, complicated and even impossible to construct circuit synthesizers for each dialect (using formal circuit representations to denote the semantics) since a single dialect even cannot construct a complete circuit. Thus, it sounds more reasonable to construct the semantics via capturing the behavior of the operations in the dialects. 

% insight from OS formalization
%% 一一解决上面提到的challenge
%% 语义方面感觉实现完HW、Comb、Seq和SV才算是完整。
\textbf{Insight \& Solution}. We present the first formal semantics of the core dialects in CIRCT, including HW, Comb, Seq, and SV dialects, since they are the most stable representations in the infrastructure and other dialects can compile to them. We employ the $\mathbb{K}$ framework to construct the semantics and generate various tools for it. We first formalize the core MLIR, including the syntax, preprocess of operations, and the builtin dialect, to provide a solid and extendable foundation for the semantics of the dialects and the consistent composition of them. Then we formalize the semantics of the CIRCT core dialects, including HW, Comb, Seq, and SV dialects. With the previous foundation, we only need to focus on the distinctive features of each dialect. Finally, we propose a benchmark and co-simulation framework to end-to-end validate the semantics, by comparing the simulation result with verilator. The benchmark includes the semantic \textbf{unit test for each operation} in the dialects to help the development of formal semantics and validate the correctness of each operation. The benchmark also includes some famous hardware designs, including Rocket, BOOM, and Reiscinator, from https://github.com/circt/arc-tests to validate the effectiveness of the semantic. (Optional BOOM, Reiscinator, and tests from verilator)

Finally, we propose a semantic unit test for each operation in the dialects to validate the formal semantics and utilize the Rocket chip to validate that the core dialects are enough to represent real-world hardware.
%% 要不要增加和documentation/ods/...的同步维持？
%% test 
%% translation Validation
%% compile
%% simulation
%% more applications
%% powerful and reliable hardware development platform

\textbf{Contributions}.

\begin{enumerate}
    \item We introduce \textbf{K-CIRCT, a framework for formalizing CIRCT dialects.} It provides the core concepts and preprocesses of MLIR semantics, which is the foundation to formalize and compose the dialect semantics. With the framework, we can focus on the semantics of each dialect without concerning the relationship and consistency.
    \item We formalize the semantics of CIRCT core dialects, including HW, COMB, Seq, SV dialects. Since other CIRCT dialects can be compiled into the core dialects, this semantics is enough for real-world hardware (Rocket) simulation and verification.
    \item We introduce a \textbf{C++ interface} to simulate the hardware, like verilator, based on the CIRCT semantics. 
    \item We propose a benchmark and co-simulation framework of CIRCT core dialects semantics. The benchmark includes the operation-level test cases for individual operations and the chip-level test cases for the whole semantics.
    % https://github.com/circt/arc-tests
    % 语义覆盖是完整；通过对比反应K的语义是对的；
    % 从性能角度，verilator是底层的仿真，我们是circt的仿真。。
    % 告诉别人实时的结果是什么样子，不一定要快。慢也是有价值的。关键是这个事实。
\end{enumerate}

\red{roadmap. Section XXX}
\red{maybe need to explain why we just focus the core dialects?}